---
title: "Project2_Khan"
author: "Zahra Kham"
format: html 
---

```{r, echo = FALSE, results='hide', warning= FALSE, message= FALSE}
library(here)
library(tidyverse)
library(RColorBrewer)



aq18 <- read.csv("AQ_2018.csv")
aq19 <- read.csv("AQ_2019.csv")
aq20 <- read.csv("AQ_2020.csv")
```

```{r, echo = FALSE, results='hide', warning= FALSE, fig.show='hide'}
# just examining what I have 
colnames(aq18)
colnames(aq19)
colnames(aq20)

#CO
boxplot(aq18$Daily.Max.8.hour.CO.Concentration) # have some outliers clearly
boxplot(aq19$Daily.Max.8.hour.CO.Concentration) # more outliers than 18
boxplot(aq20$Daily.Max.8.hour.CO.Concentration) # few but noticeable outliers

# PM
boxplot(aq18$Daily.Mean.PM2.5.Concentration) 
hist(aq18$Daily.Mean.PM2.5.Concentration) # normally dist, some negative values
hist(aq19$Daily.Mean.PM2.5.Concentration) # all pos
hist(aq20$Daily.Mean.PM2.5.Concentration) # all pos
boxplot(aq20$Daily.Mean.PM2.5.Concentration) 
# Ozone
boxplot(aq18$Daily.Max.8.hour.Ozone.Concentration)# looks good
boxplot(aq19$Daily.Max.8.hour.Ozone.Concentration) # no outliers, nice
boxplot(aq20$Daily.Max.8.hour.Ozone.Concentration) # low conc outliers

```
```{r, echo=FALSE, results='hide'}
unique(aq18$units_CO) #NA
unique(aq18$units_PM) # NA
unique(aq18$units_O3) # NA


unique(aq19$units_CO) #NA
unique(aq19$units_PM) # NA
unique(aq19$units_O3) # NA

unique(aq20$units_CO) # no NA
unique(aq20$units_PM) # NA
unique(aq20$units_O3) # no NA
```

# Background

Sickle cell disease (SCD) is a group of genetic disorders in which red blood 
cells contort into a sickle shape. Symptoms of SCD include anemia (caused by 
misshapen cells dying early) as well as pain, infection, and stroke (caused by 
sickle cells blocking blood flow). There is some evidence that exposure to air 
pollution aggravates vaso-occlusion and pain in SCD patients.

A hematologist at Duke wants to study the associations between various air 
pollutants and SCD in Durham County, North Carolina.
She provides data on health system utilization (i.e., emergency department 
visits, hospitalizations, etc.) for SCD and you obtain
air pollution data from the EPA. The pollutant data are raw measurements taken 
from the air pollution monitor at the RDU
airport. This exposure data is messy and will need to be processed. You will 
write a function to process your data and do some exploratory EDA using plots.


```{r, echo=FALSE, results='hide'}
# function part one
# I will show progress of my function and at the end will only have the finished one run
aq_process <- function(data1) {
  # want to first rename pollutant concentration cols
  data1 <- data1 |> 
    mutate(
      Dates = mdy(Date),
      CO_lvl = Daily.Max.8.hour.CO.Concentration,
      CO_units = units_CO,
      PM_lvl = Daily.Mean.PM2.5.Concentration,
      PM_units = units_PM,
      OZONE_lvl = Daily.Max.8.hour.Ozone.Concentration,
      OZONE_units = units_O3
      ) |> 
    select(
      -Date,
      -Daily.Max.8.hour.CO.Concentration,
      -units_CO,
      -Daily.Mean.PM2.5.Concentration,
      -units_PM,
      -Daily.Max.8.hour.Ozone.Concentration,
      -units_O3
      )
  
  return(data1)
}
```

```{r, echo=FALSE, results='hide'}
# testing a dataset to see if function works
aq18 <- aq_process(aq18)
aq19 <- aq_process(aq19)
head(aq18)
# testing which duplicates exist for certain days
aq18 |> 
  group_by(Dates) |> 
  summarise(count = n()) |> 
  filter(count > 1)
```
```{r, echo=FALSE, results='hide'}
aq19 |>  # we have many days in 2019 alone with duplicates
  group_by(Dates) |> 
  summarise(count = n()) |> 
  filter(count > 1)
```

```{r, echo=FALSE, results='hide'}
aq18 <- read.csv("AQ_2018.csv", na.strings = c("", "NA", "NaN")) # 479 obs
aq19 <- read.csv("AQ_2019.csv", na.strings = c("", "NA", "NaN")) # 494 obs
aq20 <- read.csv("AQ_2020.csv", na.strings = c("", "NA", "NaN")) # 179 obs
# function part two 
# reworking function to average/minimize down to one row only
aq_process1 <- function(data1) {
  # want to first rename pollutant concentration cols
  data1 <- data1 |> 
    mutate(
      Dates = mdy(data1$Date),
      CO_lvl = Daily.Max.8.hour.CO.Concentration,
      CO_units = units_CO,
      PM_lvl = Daily.Mean.PM2.5.Concentration,
      PM_units = units_PM,
      OZONE_lvl = Daily.Max.8.hour.Ozone.Concentration,
      OZONE_units = units_O3
      ) |> 
    select(
      -Date,
      -Daily.Max.8.hour.CO.Concentration,
      -units_CO,
      -Daily.Mean.PM2.5.Concentration,
      -units_PM,
      -Daily.Max.8.hour.Ozone.Concentration,
      -units_O3
      ) 
  data1 <- data1 |> 
    group_by(Dates) |> # initally treid to filter only duplicates, but reomved
    # single day observations, which is an issue
    # this will average over days with one row only but should not impact
    # the results
    summarise(
      CO_lvl    = mean(CO_lvl, na.rm = TRUE),
      CO_units  = first(CO_units),
      PM_lvl    = mean(PM_lvl, na.rm = TRUE),
      PM_units  = first(PM_units),
      OZONE_lvl = mean(OZONE_lvl, na.rm = TRUE),
      OZONE_units = first(OZONE_units),
      .groups = "drop"  # removing our duplicate rows now after taking mean
    ) |> 
    # problem with NaN sowing up after doing mean() on NAs
    # should properly turn all NaN to NA for consistency
     mutate(across(c(CO_lvl, PM_lvl, OZONE_lvl), ~na_if(., NaN))) 
  
  return(data1) 
}

```

```{r, echo=FALSE, results='hide'}
# testing 2
aq18 <- aq_process1(aq18) # 361 obs (mising four days?)
aq19 <- aq_process1(aq19) # 365 obs
aq20 <- aq_process1(aq20) # 121 obs
```

```{r, echo=FALSE, results='hide'}
aq18 |>  # no duplicates yay!
  group_by(Dates) |> 
  summarise(count = n()) |> 
  filter(count > 1)

aq19 |>  # duplicates still exist...
  group_by(Dates) |> 
  summarise(count = n()) |> 
  filter(count > 1)

# testing why aq17 has 361 observations, do not have data between 9/13-9/16 
full_dates <- seq.Date(min(aq18$Dates), max(aq18$Dates), by = "day")

missing_days <- setdiff(full_dates, unique(aq18$Dates))

missing_days
```
```{r, echo=FALSE, results='hide'}
summary(aq18$CO_lvl) # min = 0.1 max = .9; good
summary(aq18$PM_lvl) # not good, negatives
summary(aq18$OZONE_lvl) # 0.015, max = .063; good

summary(aq19$CO_lvl) # good
summary(aq19$PM_lvl) # good
summary(aq19$OZONE_lvl) # good

summary(aq20$CO_lvl) # good 
summary(aq20$PM_lvl) # good 
summary(aq20$OZONE_lvl) # good
```
```{r, echo=FALSE, results='hide'}
# function part three -> final part
# adding in for loop to remove negative values to fix measure errors
aq18 <- read.csv("AQ_2018.csv", na.strings = c("", "NA", "NaN")) # 479 obs
aq19 <- read.csv("AQ_2019.csv", na.strings = c("", "NA", "NaN")) # 494 obs
aq20 <- read.csv("AQ_2020.csv", na.strings = c("", "NA", "NaN")) # 179 obs
# function part two 
# reworking function to average/minimize down to one row only
aq_process1 <- function(data1) {
  # want to first rename pollutant concentration cols
  data1 <- data1 |> 
    mutate(
      Dates = mdy(data1$Date),
      CO_lvl = Daily.Max.8.hour.CO.Concentration,
      CO_units = units_CO,
      PM_lvl = Daily.Mean.PM2.5.Concentration,
      PM_units = units_PM,
      OZONE_lvl = Daily.Max.8.hour.Ozone.Concentration,
      OZONE_units = units_O3
      ) |> 
    select(
      -Date,
      -Daily.Max.8.hour.CO.Concentration,
      -units_CO,
      -Daily.Mean.PM2.5.Concentration,
      -units_PM,
      -Daily.Max.8.hour.Ozone.Concentration,
      -units_O3
      ) 
  data1 <- data1 |> 
    group_by(Dates) |> # initally treid to filter only duplicates, but reomved
    # single day observations, which is an issue
    # this will average over days with one row only but should not impact
    # the results
    summarise(
      CO_lvl    = mean(CO_lvl, na.rm = TRUE),
      CO_units  = first(CO_units),
      PM_lvl    = mean(PM_lvl, na.rm = TRUE),
      PM_units  = first(PM_units),
      OZONE_lvl = mean(OZONE_lvl, na.rm = TRUE),
      OZONE_units = first(OZONE_units),
      .groups = "drop"  # removing our duplicate rows now after taking mean
    ) |> 
    # problem with NaN sowing up after doing mean() on NAs
    # should properly turn all NaN to NA for consistency
     mutate(across(c(CO_lvl, PM_lvl, OZONE_lvl), ~na_if(., NaN))) 
  pollutants <- c("CO_lvl", "PM_lvl", "OZONE_lvl")
  for(i in pollutants) {
    col <- data1[[i]] # double brackets call out vector in our df
    col[col < 0] <- 0 
    data1[[i]] <- col
  }
  return(data1) 
}
```

```{r, echo=FALSE, results='hide'}
aq18 <- aq_process1(aq18) # 361 obs 
aq19 <- aq_process1(aq19) # 365 obs
aq20 <- aq_process1(aq20) # 121 obs
```


```{r, echo=FALSE, results='hide'}
# all good! all negative values removed

summary(aq18$CO_lvl) 
summary(aq18$PM_lvl)  # main issue here resolved
summary(aq18$OZONE_lvl) 

summary(aq19$CO_lvl) 
summary(aq19$PM_lvl) 
summary(aq19$OZONE_lvl) 

summary(aq20$CO_lvl) 
summary(aq20$PM_lvl) 
summary(aq20$OZONE_lvl) 
```

```{r, echo=FALSE, results='hide'}
# joining our dataset for ease of plotting in next step with rbind()
aq_full <- rbind(aq18, aq19, aq20) # 361 + 365 + 121 = 847
dim(aq_full) # joined successfully 
head(aq_full) # checking column names

# will now be using aq_full dataset from here on out
```

# Plots

```{r, warning = FALSE, echo = FALSE, results= 'hide'}
# plot 1
aq_monthly <- aq_full |>  # wanting the months on our axis only not full date
  mutate(Month = month(Dates, label = TRUE, abbr = TRUE)) |> 
  group_by(Month) |> 
  summarise(
    CO_avg = mean(CO_lvl, na.rm = TRUE), # by month we take means 
    CO_sd = sd(CO_lvl, na.rm = TRUE),
    CO_n = n() - sum(is.na(CO_lvl)), # removing NA from n 
    O3_avg = mean(OZONE_lvl, na.rm = TRUE),
    O3_sd = sd(OZONE_lvl, na.rm = TRUE),
    O3_n = n() - sum(is.na(OZONE_lvl)), # removing NA from n
    .groups = "drop" # dropping other rows we got avg from
  ) |> 
  mutate(
    CO_se = CO_sd / sqrt(CO_n),               # standard error
    O3_se = O3_sd / sqrt(O3_n),
    CO_upper = CO_avg + 1.96 * CO_se,        # 95% CI
    CO_lower = CO_avg - 1.96 * CO_se,
    O3_upper = O3_avg + 1.96 * O3_se,
    O3_lower = O3_avg - 1.96 * O3_se
  ) 
head(aq_monthly)
# now need to find  sd -> SE, to get our C.I. at 95%
```


```{r, warning = FALSE, echo = FALSE}

ggplot(aq_monthly, aes( x = Month)) + 
  geom_point(aes(y = CO_avg, color = "CO")) + # plotting points for CO first
  geom_errorbar(aes(ymin = CO_lower, ymax = CO_upper, color = "CO"), width = 0) +
  geom_point(aes(y = O3_avg, color = "O3")) + # plotting O3 now 
  geom_errorbar(aes(ymin = O3_lower, ymax = O3_upper, color = "O3"), width = 0) +
  # removed top bars as it made O3 look scrunched up and odd with small CI
  scale_color_manual(values = c("CO" = "dodgerblue2", "O3" = "darkorange1")) + 
  # chose colors that stand out best
  labs(
    title = "Plot1.Monthly Average CO and Ozone Concentrations (2018-2020)",
    y = "Concentrations",
    x = "Month",
    color = "Pollutant" # to fix legend
  ) + 
  theme_bw()

```


```{r, fig.width=10, fig.height= 12, warning = FALSE, echo = FALSE}
# plot 2 
# will copy what I did for Month from pot1 for here
pm_monthly <- aq_full |>  # wanting the months on our axis only not full date
  mutate(
    Month = month(Dates, label = TRUE, abbr = TRUE),
    Year = year(Dates),
    Years = as.factor(Year)
    ) 
         
ggplot(pm_monthly, aes(x = Month, y = PM_lvl, color = Years)) +
  geom_boxplot() +
  facet_wrap(~ Years, ncol = 1, scales = "free_x")   +
  scale_color_manual(values = c("2018" = "mediumvioletred", "2019" = "seagreen3", "2020" = "royalblue2")) +
  labs(
    title = "Plot2.PM2.5 Concentration by Month (2018-2020)",
    y = "PM2.5",
    x = "Month",
    color = "Year"
  ) + 
  theme_bw()


```


